\chapter{Introduction}

% Extend from worst-case to average-case and with additional structures.

% More question naturally arise: limiting distribution (take Quicksort)

Since Knuth's anouncement of the term \textit{analysis of algorithms} \cite[Preface]{knuth1997art}, the field has grown to include highly specialized subfields. A basic classification is based on two aspects. The first aspect is the scenario under which the performance of an algorithm is analyzed, which concludes worst-case and average-case analysis. In general, average-case analysis is more informative than worst-case analysis, and is often more challenging to perform, since it requires an probabilistic model of the input that is tractable, yet expressive enough. For example, in the case of regular languages, it may not be a good idea to generate their syntax trees uniformly at random \cite{koechlin2019uniform}. Interestingly, average-case analysis opens up the door to apply techniques from other fields, such as probability theory and analytic combinatorics, and raises new question, such as those about the limiting objects when the input size tends to infinity. The second aspect is the underlying underlying cost model, which can be the uniform or word-RAM model. The main purpose is to align quantitative analysis with practical performance, demonstrated by execution time. Moderns analysis of modern algorithms requires new cost models to natively capture hardware features, such as caching, branch prediction and parallelism.

In this intership, we study Hyperscan a regular expression matching engine that leverages SIMD parallelism to achieve high performance. Acknowledging the sophisticated design of Hyperscan, we proposed related questions and attempted to answer them. Hyperscan works by extracting possible strings from a regular expression, and then matching them once in parallel by its SIMD string matchers. Hence it is natural to analyze statistics related to those strings under a distribution. In particular, our question is to understand the asymptotic of the length of the longest strings extracted from a regular expression. Other questions are also posed clearly although not studied to a considerable extend. Details are given in the next chapter.

The rest of the report is organized as follows. In \textbf{Chapter 2}, we present Hyperscan and problems arised in the analysis. \textbf{Chapter 3} contains our results on the question of the length of the longest strings. Before each proof, we present the motivation and reasoning behind our approaches, which the reader can optionally skip and then go back after reading the proof. We place details of popular theoretical and experimental techniques in \textbf{Chapter 4}, where the reader can refer to for background knowledge. Finally, we conclude in \textbf{Chapter 5} with a summary of our work and future directions.

